---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: yes
    self_contained: no
---

## Loading and preprocessing the data
```{r}
rm(list=ls())
unlink('activity.csv')
unzip('activity.zip')
activity <- read.csv('activity.csv', header = T, sep = ',')
activity$date <- as.Date(activity$date)
```

## What is mean total number of steps taken per day?
```{r}
daily_activity <- aggregate(activity$steps, list(activity$date), FUN = sum)
names(daily_activity) <- c('date', 'Total.Steps')
hist(daily_activity$Total.Steps, 
     xlab = 'Total steps taken in a day', 
     main = 'Daily Activity', 
     breaks = 25, xlim = c(0, 25000), ylim = c(0,10))
```

#### The average of the total number of steps per day:
```{r}
mean(daily_activity$Total.Steps, na.rm = T)
````

#### The median of the total number of steps per day:
```{r}
median(daily_activity$Total.Steps, na.rm = T)
```

## What is the average daily activity pattern?
```{r}
steps_per_interval_avgdoveralldays <- aggregate(activity$steps, list(activity$interval), FUN = mean, na.rm = T)
names(steps_per_interval_avgdoveralldays) <- c('interval', 'steps')
par(mfrow = c(1,1))
plot(steps_per_interval_avgdoveralldays$interval, steps_per_interval_avgdoveralldays$steps, type = 'l', 
     xlab = '5 minute interval', 
     ylab = 'Average (over all days) number of steps per interval', 
     main = 'Daily Activity Pattern')

interval_with_max_avg_number_steps <- steps_per_interval_avgdoveralldays[
which(steps_per_interval_avgdoveralldays$steps == max(steps_per_interval_avgdoveralldays$steps, na.rm = T))
,'interval']
```
#### The interval with the maximum average numebr of steps:
```{r}
interval_with_max_avg_number_steps
```

## Imputing missing values
#### Number of missing values:
```{r}
number_of_missing_values <- sum(is.na(activity$steps))
number_of_missing_values
```

#### New dataset: Replace missing values with the mean over all days for that interval. I do this by merging in the dataset from the previous step, and replacing the missing values with the means in the merged-in data.
```{r}
activity_imputed <- merge(activity, steps_per_interval_avgdoveralldays, by.x = 'interval', by.y = 'interval')
activity_imputed <- activity_imputed[order(activity_imputed$date),]
index_of_missing_values <- is.na(activity_imputed$steps.x)
activity_imputed[index_of_missing_values, 'steps.x'] <- activity_imputed[index_of_missing_values, 'steps.y']
activity_imputed$steps.y <- NULL
names(activity_imputed) <- c("interval", "steps", "date")
head(activity_imputed)

daily_activity_imputed <- aggregate(activity_imputed$steps, list(activity_imputed$date), FUN = sum)
names(daily_activity_imputed) <- c('date', 'Total.Steps')
hist(daily_activity_imputed$Total.Steps, 
     xlab = 'Total steps taken in a day', 
     main = 'Daily Activity: using imputed values in place of missing values', 
     breaks = 25, xlim = c(0, 25000), ylim = c(0,20))
```


#### The average of the total number of steps per day (using imputed values in place of missing values):
```{r}
mean(daily_activity_imputed$Total.Steps, na.rm = T)
```

#### The median of the total number of steps per day (using imputed values in place of missing values):
```{r}
median(daily_activity_imputed$Total.Steps, na.rm = T)
```
#### Using imputed values in place of missing values did not change the mean, but changed the median to become equal to the mean.

## Are there differences in activity patterns between weekdays and weekends?
```{r}
activity_imputed$day.of.week <- weekdays(activity_imputed$date)
activity_imputed$type.of.day <- 'Weekday'
activity_imputed$type.of.day[activity_imputed$day.of.week == 'Sunday' | activity_imputed$day.of.week == 'Saturday'] <- 'Weekend'
activity_imputed$type.of.day <- as.factor(activity_imputed$type.of.day)
table(activity_imputed$type.of.day)
str(activity_imputed)
avg_over_type_of_day_number_of_steps_taken_by_interval <- aggregate(activity_imputed$steps, list(activity_imputed$type.of.day, activity_imputed$interval), FUN = mean)
names(avg_over_type_of_day_number_of_steps_taken_by_interval) <- c('type.of.day', 'interval', 'steps')
str(avg_over_type_of_day_number_of_steps_taken_by_interval)
library(lattice)
xyplot(steps~interval|factor(type.of.day), data = avg_over_type_of_day_number_of_steps_taken_by_interval,
           type = 'l',
           xlab = 'Interval',
           ylab = 'Average Number of Steps for the Interval',
           main = 'Differences in activity patterns between weekdays and weekends',
           layout=c(1,2))
unlink('activity.csv')
```
